A Mechanised Library for Generic (Co)Programming

### 2024-03-05 paper/paper.tex
Algebra.

\subsection{Algebras}
\label{sec:algebras}
An algebra is a set of abstract operations on a type \coq{X}. For example,
give such a type \coq{X} we can define the list operations on \coq{X} as
\coq{Empty : unit -> X} and \coq{Cons : A * X -> X}.

We can generalise this using containers. Given a type `\coq{A}` and a
container \coq{F} an algebra for \coq{F} over the carrier `\coq{A}` is a map of
the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}

The type of lists over a type \coq{A}, namely \coq{List A} is the ``least set''
generated by the constructors \coq{Empty : unit -> List A}
and \coq{Cons : A * List A -> List A}.  Together, \coq{Empty} and \coq{Cons}
form an \emph{algebra} of type \coq{In : unit + A * List A -> List A}.

When \coq{A} is the least fixed-point of \coq{F} this map is an isomorphism and
it is called the \emph{initial algebra} and denotes the inductive data type
shaped by the functor represented by \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
In particular, this algebra is an isomorphism with the obvious
inverse \coq{InOp : List A -> unit + A * List A} returning \coq{tt} is the list
is empty and the head and tail of the list otherwise. This type of algebra is
called \emph{initial algebra}.  This is because for for any other
algebra \coq{alg : unit + A * List B -> B} there exists a unique map which we
call \coq{cata alg : List A -> B} which destructs a list into its head and tail
using a destructor \coq{InOp : List A -> unit + A * List A}, applies itself
recursively, composes together the results of the recursion using the
algebra \coq{alg}.

The most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. Its dual is the \emph{unfold} or
\emph{anamorphisms} which produces data.

A fold consumes elements of an inductive data type. These can be thought as
least ``sets'' generated by some constructors.
% To generalise recursion schemes we abstract the shape of the data using functors.
% Assuming a base functor \coq{F} a catamorphism is then the unique
% solution to the equation
% \begin{minted}{coq}
% cata alg = alg . fmap (cata alg) . InOp
% \end{minted}
% which does precisely what we described in the previous paragraph.

@@

%%\dcas{Boring stuff}
%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
%own custom induction scheme for initial algebras:
%\begin{coqcode}
%Lemma LFix_ind [P : LFix -> Prop]
%  : (forall x : App F LFix,
%        (forall e : Pos (shape x), P (cont x e)) ->
%        P (LFix_in x))
%    -> forall l : LFix, P l.
%\end{coqcode}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \coq{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \coq{LFixR}.
\begin{coqcode}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{coqcode}

Least fixed points of container extensions come with two functions that are
defined as wrappers around constructor \coq{LFix_in} and projection
\coq{LFix_out}.
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F
l_out : LFix F ~> App F (LFix F)
\end{coqcode}

We define \emph{catamorphisms} by recursively traversing their structure of the
input \coq{LFix}. As usual, a catamorphism takes an \coq{F}-algebra with
carrier \coq{A}, and produces a function from \coq{LFix F} to \coq{A}.
\begin{coqcode}
Definition cata_f (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkApp sx (fun e => f (kx e)))
  end.
\end{coqcode}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{coqcode}
Definition cata : Alg F A ~> LFix ~> A := ...
\end{coqcode}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \coq{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \coq{f} and
\coq{cata g}.
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{coqcode}
\subsection{Coalgebras}
Terminal coalgebras are defined analogously to initial algebras (see Figure~\ref{fig:term_coalg}).
\begin{figure}
\begin{coqcode}
Context `(F : Cont Sh Po).
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
Definition GFixR : GFix -> GFix -> Prop :=...
Instance GFix_setoid : setoid GFix := ...GFixR...
Definition g_in : App F GFix ~> GFix := ...
Definition g_out : GFix ~> App F GFix := ...
\end{coqcode}
\caption{Terminal Coalgebras}\label{fig:term_coalg}
\end{figure}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode, in an analogous way to catamorphisms and initial algebras:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

@@

\subsection{(Recursive) Coalgebras}
\label{sec:rec-coalgebras}

@@

\subsection{Hylomorphisms}
\label{sec:hylo}

>>>>

\subsection{Algebras and Catamorphisms}
\label{sec:algebras} An algebra is a set \coq{X} together with some operations
on it\footnote[1]{Sometimes and algebra comes equipped with an equational
theory, e.g. the theory of monoids, but here we focus on algerbraically free
algebras}.  For example, give such a type \coq{X} we can define the monoid
operations as \coq{Empty : unit -> X} and \coq{Concat : X * X -> X}. Notice
that \coq{unit + X * X} is a functor on \coq{X} which means we can equivalently
describe this algebra as a map \coq{unit + X * X -> X} (together with the
monoidal laws).  In this work we will be using algebras to describe the
operations of a data type and so we will not be needing additional equations on
these operations.

Since in this work we will be considering only polynomial functors we need to
formalise algebras in terms of containers and containers extensions. Given a
type `\coq{A}` and a container \coq{F} an algebra for \coq{F} over the carrier
`\coq{A}` is a map of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}

Now, the \emph{easiest} way to turn any type \coq{A} into a monoid is to define
it as the least fixed-point of the functor \coq{unit + A + X * X} on \coq{X}.
This is isomorphic to the type of lists over \coq{A} hence it is the least
fixed-point for the functor \coq{unit + A * X} as well. This is the least
``set'' generated by the constructors \coq{Empty : unit -> List A}
and \coq{Cons : A * List A -> List A}. These constructors give an
algebra \coq{LFix_in : unit + A * List A -> List A} which has and inverse which
we  call \coq{LFix_out}.

We now formalise this idea using container extensions.  For a container
extension a data type is formalised as an inductive data type parametrised by
the functor represented by \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
% To generalise recursion schemes we abstract the shape of the data using functors.
% Assuming a base functor \coq{F} a catamorphism is then the unique
% solution to the equation
% \begin{minted}{coq}
% cata alg = alg . fmap (cata alg) . InOp
% \end{minted}
% which does precisely what we described in the previous paragraph.

@@

%%\dcas{Boring stuff}
%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
%own custom induction scheme for initial algebras:
%\begin{coqcode}
%Lemma LFix_ind [P : LFix -> Prop]
%  : (forall x : App F LFix,
%        (forall e : Pos (shape x), P (cont x e)) ->
%        P (LFix_in x))
%    -> forall l : LFix, P l.
%\end{coqcode}
We define smart constructors for the isomorphism of fixed points
\coq{LFix_out}.
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F
l_out : LFix F ~> App F (LFix F)
\end{coqcode}

The most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. First a fold \emph{destructs} a data
type using \coq{LFix_out}, then calls itself recursively and then folds the
results of the recursive call using an \coq{F}-algebra.
\begin{coqcode}
Definition cata_f (alg : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      alg (MkApp sx (fun e => f (kx e)))
  end.
\end{coqcode}
It is easy to show that this function is a morphism of \coq{F}-algebras. In fact,
it is possible to define it as a map of the following type:
\begin{coqcode}
Alg F A ~> LFix ~> A
\end{coqcode}

We prove that catamorphisms satisfy the universal law, that is
that there is a unique morphism of \coq{F}-algebras from the least fixed-point of \coq{F}:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}

\subsection{Coalgebras and Anamorphisms}
Terminal coalgebras are defined analogously to initial algebras (see
Figure~\ref{fig:term_coalg}).
% \begin{figure}
\begin{coqcode}
Context `(F : Cont Sh Po).
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
Definition GFixR : GFix -> GFix -> Prop :=...
Instance GFix_setoid : setoid GFix := ...GFixR...
Definition g_in : App F GFix ~> GFix := ...
Definition g_out : GFix ~> App F GFix := ...
\end{coqcode}
% \caption{Terminal Coalgebras}\label{fig:term_coalg}
% \end{figure}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode, in an analogous way to catamorphisms and initial algebras:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

@@

\mpav{Dumped here}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \coq{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \coq{LFixR}.
\begin{coqcode}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{coqcode}

\subsection{Recursive Coalgebras and Hylomorphism}
\label{sec:rec-coalgebras}

@@

\subsubsection{Hylomorphisms}
